Digging Deeper Into the State of the Practice for Domain Specific Research Software

### 2022-02-18 StateOfPractice/Papers/ICCS-2022-SOP_MethodologyPlusExamples/DigDeep.tex
Paper shortened to 14 pages

Virtual machines (VMs) are used to provide an optimal testing environments for
each package~\cite{SmithEtAl2016}. VMs are used because it is easier to start
with a fresh environment so as to not worry about existing libraries and
conflicts. Moreover, when the tests are complete the VM can be deleted, without
any impact on the host operating system. The most significant advantage of using
VMs is to level the playing field. Every software install starts from a clean
slate, which removes ``works-on-my-computer'' errors.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies} and \rqref{RQ_PainPoints}) require going beyond
the quantitative data from the measurement template. To gain the required
insight, we interview developers using a list of 20 questions
\cite{SmithEtAl2021}. The questions cover the background of the development
teams, the interviewees, and the software itself. We ask the developers how they
organize their projects. We also ask them about their understanding of the
users. Some questions focus on the current and past difficulties, and the
solutions the team has found, or will try. We also discuss the importance of,
and the current situation for, documentation. A few questions are about specific
software qualities, such as maintainability, understandability, usability, and
reproducibility. The interviews are semi-structured based on the question list;
we ask follow-up questions when necessary.  Each interview should take about 1
hour.

@@

Since each domain will have a reasonably small scope, we will be able to view
the software as a program family. Studying the common properties within a family
via a domain analysis.  For the current methodology time constraints necessitate
a shallow analysis. For each domain a table should be constructed that
distinguishes the programs under study by their variabilities.  In research
software the variabilities are often related to assumptions.
Table~\ref{tbl_features} shows the variabilities for the LBM
software~\cite{Michalski2021}.

@@

The results for the LBM example are shown below.  The majority of LBM generated
artifacts correspond to general recommendations from research software
developers.  For LBM, a union of the three categories mostly corresponds to
recommendations made by the research software community, as shown in
Table~\ref{Tbl_Guidelines}. Although the LBM community participates in most of
the practices we found listed in the general research software guidelines, some
recommended practices were not observed or rarely observed as follows: API
documentation, a roadmap, a code of conduct, programming style guide, uninstall
instructions.

@@

The running example of LBM confirms an informal process, with elements of agile
methods. The software development process is not explicitly indicated in the
artifacts for most of the packages. However, during interviews one developer
(ESPResSo) told us their non-rigorous development model is like a combination of
agile and waterfall. Employing a loosely defined process makes sense for LBM
software, given that the teams are generally small and self-contained. One of
the developers (ESPResSo) that was interviewed also noted that an ad hoc peer
review process is used to assess major changes and additions, matching the
common recommendation.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
then compare their responses to pain points identified in the
literature~\cite{WieseEtAl2019,PintoEtAl2018}.  Pain points to watch for
include: lack of development time, cross-platform compatibility, scope bloat,
lack of user feedback, dependency management, data handling concerns,
reproducibility, and software scope determination.

An example pain point noted by LBM developers is a lack of development time. A
developer of pyLBM noted that their small development team did not have enough
time to implement new features. Small development teams are common for LBM
software packages (as shown in Figure~\ref{measurement_template_image}).

>>>>

Virtual machines (VMs) are used to provide an optimal testing environments for
each package~\cite{SmithEtAl2016} because with a fresh VM there are no worries
about conflicts with existing libraries. Moreover, when the tests are complete
the VM can be deleted, without any impact on the host operating system. The most
significant advantage of using VMs is that every software install starts from a
clean slate, which removes ``works-on-my-computer'' errors.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies} and \rqref{RQ_PainPoints}) require going beyond
the quantitative data from the measurement template. To gain the required
insight, we interview developers using a list of 20 questions
\cite{SmithEtAl2021}. The questions cover the background of the development
teams, the interviewees, and the software itself. We ask the developers how they
organize their projects and about their understanding of the users. Some
questions focus on the current and past difficulties, and the solutions the team
has found, or will try. We also discuss the importance of, and the current
situation for, documentation. A few questions are about specific software
qualities, such as maintainability, usability, and reproducibility. The
interviews are semi-structured based on the question list.  Each interview
should take about 1 hour.

@@

For the current methodology time constraints necessitate a shallow domain
analysis. For each domain a table should be constructed that distinguishes the
programs under study by their variabilities.  In research software the
variabilities are often assumptions. Table~\ref{tbl_features} shows the
variabilities for the LBM software~\cite{Michalski2021}.

@@

The results for the LBM example are shown below.  The majority of LBM generated
artifacts correspond to general recommendations from research software
developers.  For LBM, a union of the three categories mostly corresponds to
general research software recommendations (Table~\ref{Tbl_Guidelines}). Areas
where LBM developers could improve include providing: API documentation, a
roadmap, a code of conduct, programming style guide, and uninstall instructions.

@@

The LBM example confirms an informal, agile-like, process.
The development process is not explicitly indicated in the artifacts
for most of the packages. However, during interviews one developer (ESPResSo)
told us their non-rigorous development model is like a combination of agile and
waterfall. Employing a loosely defined process makes sense for LBM software,
given that the teams are generally small and self-contained. One of the
developers (ESPResSo) that was interviewed also noted that an ad hoc peer review
process is used.

@@

To answer \rqref{RQ_PainPoints}, we ask developers about their pain points and
compare their responses to the literature~\cite{WieseEtAl2019,PintoEtAl2018}.
Pain points to watch for include: cross-platform compatibility, scope bloat,
lack of user feedback, dependency management, data handling concerns,
reproducibility, and software scope.

An example pain point noted for LBM is a lack of development time. A developer
of pyLBM noted that their small development team did not have enough time to
implement new features. Small development teams are common for LBM software
packages (as shown in Figure~\ref{measurement_template_image}).
